The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-25 crelle-dictionary.tex
homomorphism to morphism, a couple typos

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over $\FF_q$, then the trace of Frobenius provides a bijection between the group $G(\FF_q)^*$ of $\ell$-adic characters of $G(\FF_q)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \qcs{E} \iso \qcs{E} \boxtimes \qcs{E},%\tag{0}
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider sheaf homorphisms $\qcs{E} \to \qcs{E}'$ which are compatible with particular choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E'}$. 
{\it A priori}, these sheaf homomorphisms do not form a category, but since $G$ is connected, the existence of the isomorphism \eqref{introbox} for $\qcs{E}$ makes it unique; as a consequence, there is no impediment to categorifying the dictionary in this case. \todo{clarify}

If $G$ is a commutative algebraic group over $\FF_q$ which is not connected, however, then the isomorphism \eqref{introbox} need not be unique, so one must keep track of the choice. This leads to the category of pairs $(\qcs{E},\mu_\qcs{E})$ where $\qcs{E}$ is a rank-one local system on $G$ and $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ is a chosen isomorphism of local systems on $G\times G$. 
%If the component group scheme $\pi_0(G)$ is not cyclic, then 
In this case, the trace of Frobenius provides an epimorphism from isomorphism classes of pairs $(\qcs{E},\mu_\qcs{E})$ to characters of $G(\FF_q)$, but that group homomorphism need not be injective; consequently,
every character of $G(\Fq)$ may be geometrized as a pair $(\qcs{E},\mu_\qcs{E})$ but perhaps not uniquely in the category of pairs. 
While this might appear to be a defect in the dictionary for characters of commutative algebraic groups over finite fields, it is not serious: if $(\qcs{E},\mu_\qcs{E})$ and $(\qcs{E}',\mu_\qcs{E}')$ determine the same character of $G(\FF_q)$ then $\qcs{E}\iso \qcs{E}'$ as local systems on $G$.

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\FF_q$,
we simply replace the local system $\qcs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\qcs{E}$. 
In this way we are led to the category $\QC(G)$ of {\it quasicharacter sheaves} on $G$ (\S\ref{ssec:category}).
Objects in $\QC(G)$ are triples $(\gqcs{L}, \mu,\phi)$, where $\gqcs{L}$ is a rank-one local system on ${\bar G} := G\times_{\Spec{\FF_q}} \Spec{{\bar \FF}_q}$
and $\phi : \Frob{G}^* \gqcs{L}\to \gqcs{L}$ and $\mu : {\bar m}^* \gqcs{L} \iso \gqcs{L} \boxtimes \gqcs{L}$ are isomorphisms of sheaves satisfying a certain compatibility condition.
Morphisms in $\QC(G)$ are then homomorphisms of Weil sheaves which are compatible with the extra structure.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$ (\S\ref{sec:etale});
\'etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tools for understanding the trace of Frobenius in the \'etale case are a reinterpretation of $\QC(G)$ in terms of stalks (\S\ref{ssec:stalks}) and a
relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\QCiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that $\TrFrob{G}$ is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
We find that the group homomorphism 
\[
T_G: \QCiso{G} \to G(\FF_q)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$
determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.

>>>>

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over $\FF_q$, then the trace of Frobenius provides a bijection between the group $G(\FF_q)^*$ of $\ell$-adic characters of $G(\FF_q)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \qcs{E} \iso \qcs{E} \boxtimes \qcs{E},%\tag{0}
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider morphisms $\qcs{E} \to \qcs{E}'$ of sheaves which are compatible with particular choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E'}$. 
{\it A priori}, the composition $\qcs{E} \to \qcs{E}' \to \qcs{E}''$ of two such morphisms need not be compatible with the choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E}''$ since the choice for $\qcs{E}'$ might differ.  However, for connected $G$ the isomorphism \eqref{introbox} is unique if it exists, and there is no impediment to making the dictionary categorical.

If $G$ is a commutative algebraic group over $\FF_q$ which is not connected, then the isomorphism \eqref{introbox} need not be unique, so one must keep track of the choice. This leads to the category of pairs $(\qcs{E},\mu_\qcs{E})$ where $\qcs{E}$ is a rank-one local system on $G$ and $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ is a chosen isomorphism of local systems on $G\times G$. 
%If the component group scheme $\pi_0(G)$ is not cyclic, then 
In this case, the trace of Frobenius provides an epimorphism from isomorphism classes of pairs $(\qcs{E},\mu_\qcs{E})$ to characters of $G(\FF_q)$, but that group homomorphism need not be injective; consequently,
every character of $G(\FF_q)$ may be geometrized as a pair $(\qcs{E},\mu_\qcs{E})$ but perhaps not uniquely in the category of pairs. 
While this might appear to be a defect in the dictionary for characters of commutative algebraic groups over finite fields, it is not serious: if $(\qcs{E},\mu_\qcs{E})$ and $(\qcs{E}',\mu_\qcs{E}')$ determine the same character of $G(\FF_q)$ then $\qcs{E}\iso \qcs{E}'$ as local systems on $G$.

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\FF_q$,
we simply replace the local system $\qcs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\qcs{E}$. 
In this way we are led to the category $\QC(G)$ of {\it quasicharacter sheaves} on $G$ (\S\ref{ssec:category}).
Objects in $\QC(G)$ are triples $(\gqcs{L}, \mu,\phi)$, where $\gqcs{L}$ is a rank-one local system on ${\bar G} := G\times_{\Spec{\FF_q}} \Spec{{\bar \FF}_q}$
and $\phi : \Frob{G}^* \gqcs{L}\to \gqcs{L}$ and $\mu : {\bar m}^* \gqcs{L} \iso \gqcs{L} \boxtimes \gqcs{L}$ are isomorphisms of sheaves satisfying a certain compatibility condition.
Morphisms in $\QC(G)$ are then morphisms of Weil sheaves which are compatible with the extra structure.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$ (\S\ref{sec:etale});
\'etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tools for understanding the trace of Frobenius in the \'etale case are a reinterpretation of $\QC(G)$ in terms of stalks (\S\ref{ssec:stalks}) and a
relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\QCiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that $\TrFrob{G}$ is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
We find that the group homomorphism 
\[
T_G: \QCiso{G} \to G(\FF_q)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$
determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
